---
title: "data.table primer"
author: "Ethan Kang"
date: "February 4, 2018"
output: 
  revealjs::revealjs_presentation:
    self_contained: false
    reveal_plugins: ['chalkboard', 'menu']
    theme: sky
    highlight: pygments
    reveal_options:
      menu:
        numbers: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# data.table introduction for actuaries

## Agenda

- What is data.table?
- Why you should learn it?
- data.table features in depth

## data.table

- A package that inherits from data.frame in R
- In memory computation provides fast data processing speed

## Benefits of using data.table

- Intuitive syntax of `[i = where, j = select, by = group by, ...]` is similar to SQL `where, select, group by` statements
- Fastest in memory data manipulation tool in R
    - Include plot of benchmark?
    - Time is Money, Money is Power, Power is Pizza
- Active community on stackoverflow, so your question is most likely answered already on it

## Load packages before we start

```{r}
if (!require(pacman)) install.packages('pacman')
if (!require(CASdatasets)) install.packages("CASdatasets", repos = "http://cas.uqam.ca/pub/R/")
pacman::p_load(CASdatasets, data.table)
```

# data.table is very similar to data.frame

## Create tables
```{r}
data('usmassBI')
df <- usmassBI
dt <- data.table(df)
class(df)
class(dt)
setDT(df) # modifies df to become a data.table in place without copying
class(df)
```
- `setDT` is very useful when working with large dataset as it avoids copying. For more information, read the discussion on [stackoverflow](https://stackoverflow.com/questions/41917887/when-should-i-use-setdt-instead-of-data-table-to-create-a-data-table). 

## Looking at the data.table

```{r}
data('usmassBI')
df <- usmassBI
summary(df)
summary(dt)
```


## Row subsets

- row filters using slices `:` inside `[i, ]`
```{r}
# returns row 1 to 3 
df[1:3, ]
dt[1:3, ]
```


## Row subsets different to data.frame

- Filter based on column name - less typing than data.frame
```{r}
head(df[df$provider == 'A', ], 5)
head(dt[provider == 'A', ], 5)
```

## Row subsets different to data.frame - Cont.

- Using secondary index `on` notation - this is faster than normal subset above
    - Read more about secondary index in the [vignette](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-secondary-indices-and-auto-indexing.html)
```{r}
head(dt['A', on = 'provider'], 5)
```


## Column selects

- Can use names
- Can use numbers

## Column selects unique to data.table

- Can use varaible of names
- Can use list

# How to add columns to a data.table

## Add columns

- Based on existing columns
- From vectors
- Summarize (mean, median, max, etc) with group by

# How to make data.table even faster

## Update by reference

- `:=` notation
- set operation

# How to join tables together using data.table

## Joins

- merge
- `[` notation

## Append multiple data.frames together (`Union`)

# How to do a SQL `transform` (and back) in data.table 

## Pivot table from long to wide

- dcast

## Pivot table from wide to long

- melt

# Work with data.table using variables

## Passing variables to data.table

- Inside a function using get/mget
- Applying functions to multiple columns (`.SD, .SDcols`)
