---
title: "data.table primer"
author: "Ethan Kang"
date: "February 4, 2018"
output: 
  revealjs::revealjs_presentation:
    self_contained: false
    reveal_plugins: ['chalkboard', 'menu']
    theme: sky
    highlight: pygments
    reveal_options:
      menu:
        numbers: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# data.table introduction for actuaries

## Agenda

- What is data.table?
- Why you should learn it?
- data.table features in depth

## data.table

- A package that does tablular data manipulation
- data.table inherits from data.frame in R
- In memory computation provides fast data processing speed

## Benefits of using data.table

- Intuitive syntax of `[i = where, j = select, by = group by, ...]` is similar to SQL `where, select, group by` statements
- Fastest in memory data manipulation tool in R
    - Include plot of benchmark?
    - Time is Money, Money is Power, Power is Pizza
- Active community on StackOverflow, so your question is most likely answered already on it

## Load packages before we start
Note, this next code block will install the `data.table` package if it is not already installed.

```{r}
if (!require(data.table)) install.packages("data.table")
library(data.table)
```

# data.table is very similar to data.frame

## Create tables and read data
This block creates a list of approximately one million auto claims allocated to 1966--2015, so all are assumed at ultimate. To test the speed of reading files into R, the file has already been created and saved as `SampleAutoLosses.csv`.
```{r CreateTable, eval=FALSE}
set.seed(354)
years <- seq(1966, 2015)
counts <- rnbinom(length(years), size = 3000, mu = 20000)
allCounts <- sum(counts)
smallCounts <- floor(0.7 * allCounts)
bigCounts <- allCounts - smallCounts
numPassBig <- sample(seq_len(6), bigCounts, replace = TRUE,
                     prob = c(0.4, 0.2, 0.15, 0.1, 0.08, 0.07))
numPassSmall <- sample(seq_len(3), smallCounts, replace = TRUE,
                       prob = c(0.7, 0.25, .05))
smallLosses <- rgamma(smallCounts, shape = 4, scale = 2500)
bigLosses <- rgamma(bigCounts, shape = 0.5, scale = 1e6)
sLosses_df <- data.frame(Loss = smallLosses, Pass = numPassSmall)
bLosses_df <- data.frame(Loss = bigLosses, Pass = numPassBig)
aLosses_df <- rbind(sLosses_df, bLosses_df)
losses_df <- aLosses_df[sample(nrow(aLosses_df)), ]
row.names(losses_df) <- NULL
ALAE_P = rlnorm(allCounts, meanlog = log(0.1) - log(5) / 2,
                sdlog = sqrt(log(5)))
ALAE = losses_df$Loss * ALAE_P
writeStates <- state.abb[state.division %in%
                           c('New England', 'Middle Atlantic', 'South Atlantic')]
losses_df <- data.frame(Year = rep(years, counts), Loss = round(losses_df$Loss, 0),
                        ALAE = round(ALAE, 0), Pass = losses_df$Pass,
                        State = sample(writeStates, allCounts, replace = TRUE))
rm(years, counts, allCounts, smallCounts, bigCounts, numPassBig, numPassSmall,
   smallLosses, bigLosses, sLosses_df, bLosses_df, aLosses_df, ALAE_P, ALAE,
   writeStates)
```
```{r ReadTables}
colClass <- c('integer', rep('numeric', 2), 'integer', 'character')
df_T <- system.time(df <- read.csv('./SampleAutoLosses.csv', header = TRUE,
                             colClasses = colClass))
dt_T <- system.time(dt <- fread('./SampleAutoLosses.csv', header = TRUE,
                             colClasses = colClass))
class(df)
class(dt)
setDT(df) # modifies df to become a data.table in place without copying
class(df)
```
- `setDT` is very useful when working with large dataset as it avoids copying. For more information, read the discussion on [stackoverflow](https://stackoverflow.com/questions/41917887/when-should-i-use-setdt-instead-of-data-table-to-create-a-data-table). 

## Looking at the data.table

```{r}
data('usmassBI')
df <- usmassBI
summary(df)
summary(dt)
```


## Row subsets

- row filters using slices `:` inside `[i, ]`
```{r}
# returns row 1 to 3 
df[1:3, ]
dt[1:3, ]
```


## Row subsets different to data.frame

- Filter based on column name - less typing than data.frame
```{r}
head(df[df$provider == 'A', ], 5)
head(dt[provider == 'A', ], 5)
```

## Row subsets different to data.frame - Cont.

- Using secondary index `on` notation - this is faster than normal subset above
    - Read more about secondary index in the [vignette](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-secondary-indices-and-auto-indexing.html)
```{r}
head(dt['A', on = 'provider'], 5)
```


## Column selects - using names

- Notice that `df` returns a vector instead of data.frame
```{r}
head(df[, 'claims'], 5)
```
- `dt` returns a single column data.table, `data.table` tries to be consistent in returning data.table object from user operation
```{r}
head(dt[, 'claims'], 5)
```
## Column selects - using names - Cont.

- To return a vector from a data.table, one can use the following
```{r}
head(df[['claims']], 5)
head(df$claims, 5)
head(dt[['claims']], 5)
head(dt$claims, 5)
```

## Column selects - multiple names

```{r}
head(df[, c('claims', 'provider')], 5)
head(dt[, c('claims', 'provider')], 5)
```


## Column selects - using numbers

- Although selecting by column number is not good practice in general, it is supported to help data.frame users transition more smoothly to data.table
```{r}
head(df[, 1:3], 5)
head(dt[, 1:3], 5)
```

## Column selects different to data.frame - use list

- `.()` is an alias for `list()`, this shortens the syntax
- You don't need to have quotation mark around the column names
```{r}
head(dt[, list(claims, provider)], 5)
head(dt[, .(claims, provider)], 5)
```


## Column selects different to data.table - variable approach with `mget` or `..`

- This is useful when you have several columns you want to select and don't want to clutter your data.table operation
- Or you are passing variables in a function call
```{r}
cols_to_select <- c('claims', 'provider', 'providerA', 'logclaims')
head(df[, cols_to_select], 5)
#head(dt[, cols_to_select], 5) # because you can call column name directly, data.table will try to look for cols_to_select as a column in the table
# Error in `[.data.table`(dt, , cols_to_select) : j...
```
## Column selects different to data.table - variable approach with `mget` or `..` - Cont.

- Use `mget` to get the column names in the correct scope
```{r}
head(dt[, mget(cols_to_select)], 5)
```
- Use `..` to achieve the same result
```{r}
head(dt[, ..cols_to_select], 5)
```


# How to add columns to a data.table

## Add columns

- Based on existing columns
- From vectors
- Summarize (mean, median, max, etc) with group by

# How to make data.table even faster

## Update by reference

- `:=` notation
- set operation

# How to join tables together using data.table

## Joins

- merge
- `[` notation

## Append multiple data.frames together (`Union`)

# How to do a SQL `transform` (and back) in data.table 

## Pivot table from long to wide

- dcast

## Pivot table from wide to long

- melt

# Work with data.table using variables

## Passing variables to data.table

- Inside a function using get/mget
- Applying functions to multiple columns (`.SD, .SDcols`)
